---
title: "Benchmark Data Manipulation"
author: "Shian Su"
date: "`r Sys.Date()`"
output: BiocStyle::pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(CellBench)
```

# Introduction

This vignette contains some examples of common manipulations of data objects in this package. This package is built around the tidy data ideas established by Hadley Wickham's [tidyverse](https://www.tidyverse.org), the primary goals are to keep data in an organised manner and enable concise manipulations for achieving a wide range of outcomes.

# Benchmark Tibble

## Basics

The fundamental object in this package is the benchmark tibble. [Tibbles](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html) are variant of the `data.frame`, they are used here for their nicer printing properties.

The benchmark tibble is a `data.frame` structure where the `result` column is a special _list-column_ which is allowed to contain arbitrary data types compared to a regular column which may only contain [atomic data types](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Vector-objects) such as `numeric`, `logical` and `character`. A _list-column_ is a list with the same number of elements as there are rows in the data frame, they exist because the typical column is a vector which cannot contain complicated data types. However if the results of a computation are simple atomic values, then the column will be coerced to a regular vector with all the expected behaviours.

**IMPORTANT!** Because the result column is a list, care must be taken when performing certain operations. Most vectorised operations do not work on lists, and when we will cover how to properly work with these columns in the [Operations On Benchmark Tibbles](#ops-on-benchmark-tibbles) section.

We demonstrate how the benchmark tibble works:

```{r}
library(CellBench)
datasets <- list(
    random_mat1 = matrix(runif(100), 10, 10),
    random_mat2 = matrix(runif(100), 10, 10)
)

cor_method <- list(
    pearson = function(x) cor(x, method = "pearson"),
    kendall = function(x) cor(x, method = "kendall")
)

res <- datasets %>%
    apply_methods(cor_method)
```

As we can see, the table contains the data used, methods applied and the result of the computation. The reason for using tibbles is so that the result column is printed in a summarised form rather than fully expanded as would be the case for non-tibble list-columns.

## Operations On Benchmark Tibbles {#ops-on-benchmark-tibbles}

```{r}
class(res)
```

The benchmark tibble inherits from tibbles which inherit from data.frame, so operations expected to work on the parent classes should be expected to work on the benchmark tibble.

```{r}
res[1:2, ]
```

By default tibbles only print the first 10 rows, this doesn't change with how many elements you subset. Instead you should use `print(res, n = Inf)` if you wish to print the whole tibble, or a desired number of rows.

We can also make use of the `dplyr` functions along with piping to write concise expressions for manipulating the benchmark tibble.

```{r}
library(dplyr)
res %>%
    filter(cor_method == "pearson")
```

It is also possible to cbind two benchmark tibbles together, for example if you had added another set of methods 

```{r}
cor_method <- list(
    spearman = function(x) cor(x, method = "spearman")
)

res2 <- datasets %>%
    apply_methods(cor_method)

res2
```

```{r}
rbind(res, res2)
```

This allows new methods to be added without having to recompute results for old methods.

## Operations On list-columns

We note again that the benchmark column is a list.

```{r}
class(res$result)
```

This means some simple vectorised functions will not quite work as expected. For example if we wished to take the exponential of all the matrices using `dplyr::mutate`. Because `dplyr` feeds entire columns into the functions and expects the entire column to be returned, the result of the following code will attempt to run `exp()` on a `list` which it cannot handle.

```{r, eval = FALSE}
# this code will fail
res %>%
    mutate(exp_result = exp(result))
```

Instead we must reformulate these to expressions that take in list arguments and return lists or vectors of the same length. This can be done using either `lapply` from the base R library or `map` from the `purrr` package.

```{r}
res %>%
    mutate(exp_result = lapply(result, exp)) %>%
    mutate(sum_of_exp = unlist(lapply(exp_result, sum)))
```

## Unnesting with Lists of data.frames

One of the most useful representations that can be created in the tibble framework is to have data frames with consistent columns as the `result` list-column. This allows the data to be unnested such that the contents of the result data frames are row-contenated and the information in the remaining rows are duplicated accordingly.

# Manipulating Functions

## Basics of Functional Programming

## Partial Application

## Memoisation

# Further Reading

* Introduction to tibbles: https://tibble.tidyverse.org
* Introduction to purrr: https://purrr.tidyverse.org

